# Comparing airlines using a tidied flight delay data set

## Data 607 assignment

### Heather Geiger ; February 28, 2018

### Preface - Comparing different ways to tidy and transform data

Inspired by this blog post (https://www.r-bloggers.com/how-to-reshape-data-in-r-tidyr-vs-reshape2/), where appropriate I will show how to perform the same operation using two different functions, either reshape2 or base R versus the tidyverse tidyr/dplyr.

My main experience is using base R, with a bit of reshape2 for simple operations like switching between wide and long. Here I typically first looked into how to perform the function using my existing knowledge, before also searching for the equivalent in the tidyverse.

### Load libraries.

Load all libraries we will need.

```{r}
library(stringr)
library(reshape2)
library(tidyr)
library(dplyr)
library(ggplot2)
library(scales)
```

### Initial processing

Start by reading in the CSV file.

This CSV file was created by putting the information into Excel exactly as it looked in the description, then converting this sheet to CSV.

```{r, echo=TRUE,eval=TRUE}
airlines <- read.csv("https://raw.githubusercontent.com/heathergeiger/Tidying_airlines_data/master/alaska_vs_am_west_airlines.csv",check.names=FALSE,stringsAsFactors=FALSE)
airlines
```

Remove empty lines and make column headers more descriptive.

```{r}
airlines <- airlines[which(airlines[,2] == "on time" | airlines[,2] == "delayed"),]
colnames(airlines)[1:2] <- c("Airline","Status")
airlines
```

Right now some columns are going to be character that should be numeric, because numbers greater than 1,000 are written with a comma.

Remove the commas using str_replace from the stringr package.

Then apply as.numeric function to convert all the city columns to numeric.

```{r, echo=TRUE,eval=TRUE}
for(i in 3:ncol(airlines))
{
airlines[,i] <- str_replace(airlines[,i],pattern=",",replacement="")
airlines[,i] <- as.numeric(airlines[,i])
}

airlines
```

Check the sum of one of the columns that formerly had a comma in number to be sure this worked.

```{r}
sum(airlines[,4])
221 + 12 + 4840 + 415
```

Looks like removing the commas, then applying as.numeric, worked well.

### Tidying data

Before we can convert to long format, we need to repeat the value in Airline column that is there for on time, immediately after for delayed.

```{r}
airlines$Airline <- rep(airlines$Airline[which(airlines$Airline != "")],each=2)
airlines
```

Now ready to convert to long format.

One way to convert from wide to long format is using function melt from the reshape2 package.

Another is using the gather function.

```{r, echo=TRUE, eval=TRUE}
melt_result <- melt(airlines,variable.name="City",value.name="Flights",id.vars=c("Airline","Status"))
melt_result

gather_result <- gather(airlines,City,Flights,-Airline,-Status)
gather_result
```

These are identical. Since this assignment is supposed to be based on tidyverse, let's use the gather result.

```{r}
airlines <- gather_result
```

This is mostly what we want, except we really want two columns for on time vs. delayed for each instance of airline + city.

For this, we could use either the spread function from tidyr or the dcast function from reshape2.

One we show that they make the same result, reassign airlines to the result of spread.

```{r}
dcast_result <- dcast(airlines,formula = Airline + City ~ Status,value.var="Flights")
dcast_result

spread_result <- spread(airlines,Status,Flights)
spread_result

airlines <- spread_result
```

One last thing - change "on time" to "on.time" to be more R-friendly.

```{r}
colnames(airlines)[4] <- "on.time"
```

### Transforming data

Now our data is pretty tidy, but we need to do some transformations to get anything meaningful.

Mainly for each airline in each city, we'll want to get the rate of delays and the total number of flights.

Add this information to airlines using either transform (base R) or mutate (dplyr).

```{r}
transform_result <-  transform(airlines,total.flights = on.time + delayed,delayed.rate = (delayed*100)/(on.time + delayed))
transform_result

mutate_result <- mutate(airlines,total.flights = on.time + delayed,delayed.rate = (delayed*100)/(on.time + delayed))
mutate_result
```

In this case, the syntax and result are identical.

However, the tidyverse can also be done via piping, which can be useful when we have more complex operations to perform.

Let's show this here. Then, reassign airlines to the result of mutate.

```{r}
mutate_via_pipe_result <- airlines %>% mutate(total.flights = on.time + delayed,delayed.rate = (delayed*100)/(on.time + delayed))
mutate_via_pipe_result

airlines <- mutate_result
```

### Analysis

Let's make a side-by-side barplot where we plot the delay rate by city, putting the two airlines side-by-side.

```{r}
ggplot(airlines,aes(x = City,y = delayed.rate,fill=Airline)) +
geom_bar(stat='identity',position='dodge') + 
ylab("Percent of flights delayed") +
theme(axis.title.x=element_blank(),axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
```

Based on this, it seems pretty clear that Alaska airlines performs better, having a lower delay rate in every city.

### Plotting the same in base R

Just for fun, let's try to make a similar plot in base R.

We'll have to convert the data back to wide format first. Then, use barplot function from base R.

```{r}
delayed_rates_wide <- spread(airlines[,c("Airline","City","delayed.rate")],City,delayed.rate)
rownames(delayed_rates_wide) <- delayed_rates_wide$Airline
delayed_rates_wide <- delayed_rates_wide[,2:ncol(delayed_rates_wide)]

barplot(as.matrix(delayed_rates_wide),beside=T,las=2,ylab="Percent of flights delayed",col=c("red","blue"))
legend("topleft",legend=rownames(delayed_rates_wide),col=c("red","blue"),pch=15,bty="n")
```

That actually wasn't so bad. The major issue here though, is that the column names end up not fitting on the page using base R with default settings.

I'm sure there is a way to get them to fit better by fiddling with the margins, but in this case it's easier to just use ggplot2.

### Analysis using combined data across cities

In this case, the conclusion is pretty clear when looking city-by-city.

But what if we didn't have the city-by-city data, but only aggregate data by airline?

Use the aggregate function from base R or group_by and summarize_all from dplyr to sum delayed and on time per airline.

```{r}
aggregate_result <- aggregate(.~Airline,FUN=sum,data=airlines[,c("Airline","delayed","on.time")])
aggregate_result

summarize_result <- airlines[,c("Airline","delayed","on.time")] %>% group_by(Airline) %>% summarize_all(funs(sum))
summarize_result
```

Now, get a delayed rate similar to how we did previously.

Let's get fancy and pipe the result of summarize right into mutate, without using object summarize_result.

```{r}
airlines[,c("Airline","delayed","on.time")] %>% group_by(Airline) %>% summarize_all(funs(sum))  %>% mutate(delayed.rate = delayed*100/(delayed + on.time))
```

Interesting! If we do not separate by city, AM WEST will appear to be better in terms of delay rate.

Let's make an explanatory plot to show how this might occur.

Make a stacked barplot with the percent of flights coming from each city per airline.

Combined with the plot above, which shows trends of delay rates by city, we may get a better sense of how this happened.

```{r}
ggplot(airlines,aes(x = Airline,y=total.flights,fill=City)) +
geom_bar(position = "fill",stat = "identity") +
scale_y_continuous(labels = percent_format()) +
ylab("Percent of flights coming from each city") +
scale_fill_manual(values=c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2")) #Use colorblind vector as suggested here: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
```

We find that AM WEST has the majority of flights coming from Phoenix. And we saw from the first chart that Phoenix tends to be a lower-delay city regardless of airline.

So if we do not separate by city, AM WEST will appear to have a lower delay rate than Alaska. Whereas if we normalize by city, we come to the correct conclusion that Alaska performs better in terms of delays.
